Compiling Python Modules to Native Parallel Modules Using Pythran and OpenMP Annotations

### 2013-10-26 doc/papers/sc2013/paper.tex
SC2013: takes into account reviewer 1 comments

Pythran~\cite{pythran2013} is a subset of Python designed for scientific
computing. It is implicitly statically typed and supports most Python constructs
except those that involve introspection (e.g.\ \texttt{getattr}) or runtime
compilation (e.g.\ \texttt{eval}). A few standard modules are supported in
addition to the core language (e.g.\ \texttt{math}, \texttt{random}). The
\texttt{numpy} module is currently only partially supported. User classes are
not supported. The associated compiler turns Pythran code, possibly annotated
with OpenMP and a few function type annotations, into C++ code.
Figure~\ref{fig:pythran-compiler} summarizes its processing: type inference
helps removing most dynamic behavior, and is combined with a high-level runtime
library ---\texttt{pythonic++}--- to allow a one-to-one mapping between Python
and C++ constructs. The purpose of this section is not to dig into the internals
of the compiler, but rather to focus on the impact of the parallelism layer,
especially on the runtime library.

@@

To achieve this goal, a translator from a subset of the Python language to C++
has been designed. This translator turns regular Python modules annotated with
OpenMP directives and a few type annotations into native parallel module. The
input module remains compatible with the standard interpreter and the underlying
runtime is compatible with parallel constructs.

>>>>

Pythran~\cite{pythran2013} is a subset of Python designed for scientific
computing. It is implicitly statically typed and supports most Python constructs
except those that involve introspection (e.g.\ \texttt{getattr}) or runtime
compilation (e.g.\ \texttt{eval}). A few standard modules are supported in
addition to the core language (e.g.\ \texttt{math}, \texttt{random}). The
\texttt{numpy} module is currently only partially supported. User classes are
not supported. The associated compiler turns Pythran code, possibly annotated
with OpenMP and a few function type annotations, into C++ code.

Figure~\ref{fig:pythran-compiler} summarizes its processing: type inference
helps removing most dynamic behavior, and is combined with a high-level runtime
library ---\texttt{Pythonic++}--- to allow a one-to-one mapping between Python
and C++ constructs. Pythonic++ is a template library, i.e.\ it consists
exclusively of a set of headers. Therefore the generated C++ file is easily
distribuable and does not depend on Pythran. One could even preprocess this
C++ file to avoid the need of distributing along the Pythonic++ headers.

The purpose of this section is not to dig into the internals
of the compiler, but rather to focus on the impact of the parallelism layer,
especially on the runtime library.

@@

To achieve this goal, we designed Pythran, a translator from a subset of the
Python language to C++. Pythran turns regular Python modules annotated with
OpenMP directives and a few type annotations into native parallel module. The
input module remains compatible with the standard interpreter and the underlying
runtime is compatible with parallel constructs. While Pythran accepts a limited
input, we extend it progressively and with a great care on its stability. As
such, even if its development is community-based, it is more than a research
prototype and should be usable on real world code.
